Entrainment of low Mach number thermals in stratified domains

### 2019-05-10 tex/stratified_thermals.tex
huge changes to paper text

@@

\subsection{Theoretical description of thermal evolution}
The evolution of thermals as buoyant vortex rings has been well described in
the unstratified, Boussinesq limit for decades (as early as e.g., CITE, and
see \citet{lecoanet&jeevanjee2018} for other sources). Buoyant motions 
in the atmospheres of stars and planets are generally large enough to feel the
atmospheric stratification, and therefore a more thorough treatment of the
evolution of thermals in stratified domains is required to understand 
thermal entrainment in nature.

In this work, we focus on the non-dissipative, low Mach number regime, 
in which the ideal anelastic equations are a decent approximation to the fully
compressible equations. In this regime, buoyant forces can be expressed explicitly
in terms of the entropy. In the the limit where diffusivities
are sufficiently small and therea re no thermal sources or sinks, 
the volume-integrated total entropy inside of a buoyant vortex ring is constant,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho\, s\, dV = \text{const},
\end{equation}
where $s = c_V \ln T - R \ln\rho$ is the specific entropy, with $c_V$ the
specific heat at constant volume and $R$ the ideal gas constant.

In a stratified domain, the hydrodynamic impulse is defined
\citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector, $\bm{u}$ is vector velocity, $\rho$ is density,
and $\mathcal{V}$ is the volume being integrated over. Impulse is equivalent to the time-integrated
work acting on the fluid resulting in the current fluid motion. 
Changes in the impulse can be expressed as
\begin{equation}
\frac{\partial\bm{I}}{\partial t} = \int_{\mathcal{V}}\frac{\partial}{\partial t}(\rho \bm{u})dV
= B\hat{z} + S,
\end{equation}
where $S$ is a combination of surface terms that disappears upon appropriate boundary conditions
on $\mathcal{V}$ \citep{shivamoggi2010}. In this context where buoyancy is the only source of
work and the buoyancy, $B$, is constant, the impulse can be straightforwardly integrated as
\begin{equation}
I_z = B t + I_0,
\end{equation}
for some constant $I_0$. In the limiting regime where all of the buoyant fluid initially
located in the thermal remains in the buoyant vortex ring, $I_0 \rightarrow 0$, but as
some of the buoyant fluid is generally detrained in the initial spin-up to a vortex ring,
it is important to keep the factor of $I_0$.
Under the assumption that the thermal develops into a thin-core 
propagating vortex ring whose vortex core is radius $r$ away from the axis of symmetry, 
the impulse can be well approximated as $I_z \approx \pi \rho r^2 \Gamma$, where 
$\Gamma = \iint (\grad\times\bm{u}) dr dz$ is the circulation in the vortex ring.
Under this approximation, it becomes clear that so long as the vortex core is bounded
by $\mathcal{V}$, the volume integrated over is rather arbitrary, as most of the circulation
is located in the concentrated vortex ring.

The vertical momentum of the thermal is less straightforward to find than the momentum. 
Upon weighting the momentum equation (Eqn.~\ref{eqn:AN_momentum_full}) by the density, 
Assuming an adiabatically stratified atmosphere in
hydrostatic equilibrium, the change in volume integrated momentum over time is
\begin{equation}
\frac{D \bm{P}}{Dt} = \frac{D}{Dt}\int_{\mathcal{V}}\rho\bm{u} dV
= \int_{\mathcal{V}} \rho\frac{D\bm{u}}{Dt} dV
\approx B\hat{z} - \int_{\mathcal{V}} \rho\grad\varpi dV.
\end{equation}
As noted by \citet{tarshish&all2018}, only infinitesimally small buoyant regions of
fluid accelerate according exactly according to their buoyant nature. Any buoyant anomaly
with a finite size disrupts its fluid environment, and this disruption results in
pressure terms which reduce the efficacy of the buoyant fluid in driving motions. 
In their work with Boussinesq thermals, they find that the true buoyant acceleration
can be well approximated by scaling down the buoyant force by a factor of O($1/2$).
Taking their work into account, we assume that the vertical momentum of the thermal
follows
\begin{equation}
P_z = \beta B t + P_0,
\end{equation}
where in this work we set $\beta = 1/2$.

Rearranging, we find our first result,
\begin{equation}
r = \sqrt{\frac{B t + I_0}{\pi\rho\Gamma}}.
\label{eqn:r_theory}
\end{equation}
In the limit where $\rho \rightarrow \text{constant}$, as in the Boussinesq regime,
we retrieve the $r \propto \sqrt{t}$ scaling found in the Boussinesq regime by
\citet{lecoanet&jeevanjee2018}. We find that the inclusion of stratification adds the
additional complexity of $r \propto \rho^{-1/2}$, such that downward-propogating 
vortex rings (as studied here) will not entrain to the same degree as boussinesq thermals,
and upward-propogating rings will entrain more. In the absence of buoyancy ($B = 0$),
this result aligns with the prediction for purely horizontal compression noted by
\citet{brandenburg2016} of $r^2 \propto \rho^{-1}$.

We further assume that the volume-integrated momentum, $P_z \approx \rho V w_{th}$,
where $w_{th}$ is the vertical velocity of the thermal as a whole, and the volume
of the fluid region propogating with the thermal is a spheroid, $V = V_0 r^3$, for
some constant $V_0$. Plugging Eqn.~\ref{eqn:r_theory} into this approximation, we
find that
\begin{equation}
\rho^{-1/2}w = \left(\frac{(\pi\Gamma)^{3/2}}{V_0}\right)\frac{\beta B t + M_0}{(B t + I_0)^{3/2}}.
\end{equation}
Substituting $w = dz/dt$ and $\tau \equiv (Bt + I_0)/\Gamma$, We retrieve
\begin{equation}
\frac{dz}{\sqrt{\rho(z)}} 
= \left(\frac{\pi^{3/2}\Gamma}{V_0 B}\right)
\left(\beta\Gamma\tau^{-1/2} + [M_0 - \beta I_0]\tau^{-3/2}\right)d\tau.
\label{eqn:general_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this
result can be straightforwardly integrated with $\rho(z)$ plugged in in order to
find the position of the thermal versus time. While we will leave this result
general for now, we will plug in our polytropic stratification at the end of
section \ref{sec:experiment}, and show that the resulting expression does a
good job of explaining the evolution of thermals in these atmospheres in
section \ref{sec:results}.

\subsection{Parameterization}

\begin{deluxetable*}{c c c c c c}
\tabletypesize{\footnotesize}
\caption{Simulation output parameterization
\label{table:simulation_info}
}
\tablehead{																																															
\colhead{$n_\rho$} & \colhead{$L_{thermal}/H_\rho$} & \colhead{nr or nx = ny} & \colhead{nz} & \colhead{$t_{evolution}$} & \colhead{safety}	}	
\startdata																																															
\multicolumn{3}{l}{\textbf{2D Anelastic Simulations}}\\
0.1 	& 					&	256			& 512			& 49 	&	0.2	\\
0.5 	& 					&	256			& 512			& 42.5 	&	0.2	\\
1	 	& 					&	256			& 512			& 38 	&	0.2	\\
2	 	& 					&	256			& 512			& 31.75	&	0.2	\\
3	 	& 					&	512			& 1024			& 27.25	&	0.2	\\
4	 	& 					&	1024		& 1024			& 25 	&	0.2	\\
5	 	& 					&			& 			&  	&		\\
6	 	& 					&			& 			&  	&		\\
7	 	& 					&			& 			&  	&		\\
\multicolumn{3}{l}{\textbf{3D Fully Compressible Simulations}}\\
0.5 	& 					&	256			& 512			& 45 	&	0.2	\\
1	 	& 					&	256			& 512			& 45 	&	0.2	\\
2	 	& 					&	256			& 1024			& 45 	&	0.2	\\
3	 	& 					&	256			& 2048			& 45 	&	0.2	\\
\enddata																																															
\tablecomments{
}
\end{deluxetable*}

@@

\subsection{Anelastic Simulations}
In this work, we primarily study the evolution of 2D, azimuthally symmetric, anelastic
thermals in cylindrical coordinates. 
We later verify that these simulations produce the same results a 3D fully compressible
simulations in 3D cartesian domains (see sec. REF). 
The LBR anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\td{\grad}\cdot\td{\bm{u}} = -\td{w}\partial_{\tilde{z}} \ln\rho_0 
\label{eqn:AN_continuity_full}\\
\frac{D \td{\bm{u}}}{D \td{t}} = -\td{\grad} \td{\varpi} + g\frac{\td{S_1}}{c_P}\hat{z} + \frac{1}{\rho_0}\td{\grad}\cdot\left(\mu\td{\lilstressT}\right)
\label{eqn:AN_momentum_full}\\
\frac{D \td{S_1}}{D\td{t}} = \frac{1}{\rho c_P}\td{\grad}\cdot\left(\kappa T_0 \td{\grad} \td{S_1}\right) + \frac{\mu}{\rho_0 T_0}\td{\sigma_{ij}}\partial_{\td{x_i}}\td{u_j}
\label{eqn:AN_energy_full}.
\end{gather}
where terms with tildes are dimensional, and where
$D/D\td{t} = \partial/\partial \td{t} + \td{\bm{u}}\cdot\td{\grad}$. In our azimuthally symetric
domain, we assume that $\partial_\phi = 0$; as the initial conditions of our simulations are at rest and have
no azimuthal velocity, $u_\phi$, we assume that $u_\phi = 0$; therefore $\bm{u} = u_r \hat{r} + w\hat{z}$. 
Under this approximation, the components of the stress tensor are
\begin{equation}
\begin{split}
&\td{\sigma}_{rr} = 2\frac{\partial \td{u_r}}{\partial \td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,
\td{\sigma}_{rz}     = \td{\sigma}_{zr} = \frac{\partial \td{w}}{\partial \td{r}} + \frac{\partial \td{u_r}}{\partial \td{z}}, \\
&\td{\sigma}_{\phi\phi} = 2\frac{\td{u_r}}{\td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,\,
\td{\sigma}_{r\phi}  = \td{\sigma}_{\phi r}  = 0, \\
&\td{\sigma}_{zz}       = 2\frac{\partial \td{w}}{\partial \td{z}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}}, \,\,\,\,\,\,
\td{\sigma}_{\phi z} = \td{\sigma}_{z \phi}  = 0,\qquad
\end{split}
\end{equation}
Furthermore, we assume that the dynamic viscosity, $\mu = \rho \nu$, and the
thermal conductivity, $\kappa = \rho \chi$, are both uniform throughout the domain.
The diffusivities
$\nu$ and $\chi$ therefore scale inversely with the density.

We nondimensionalize these equations in the same manner as in
\citet{lecoanet&jeevanjee2018} such that
the length scale is the diameter of the initial thermal perturbation
and the velocity scale is the freefall velocity. The timescale is thus
the freefall crossing time of this unit length. Mathematically,
\begin{equation}
\begin{split}
\td{\grad}\rightarrow(\td{L}_{th}^{-1})\grad, \qquad&
\td{S}_1 \rightarrow(\Delta\td{S})S_1,\\
\td{\bm{u}} \rightarrow (\td{u}_{th})\bm{u}, \qquad&
\td{\varpi} \rightarrow (\td{u}_{th}^2)\varpi,\\
\partial_{\tilde{t}} \rightarrow (\td{u}_{th}/\td{L}_{th})\partial_t,\qquad&
\end{split}
\end{equation}
with
\begin{equation}
\tilde{u}_{th}^2 = \frac{g \tilde{L}_{th} \Delta \tilde{s}}{c_P}, \qquad
\text{Re}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\nu}, \qquad
\text{Pr}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\chi}.
\end{equation}
As the diffusivities scale with depth, Re$_{\text{ff}}$ is specified at the
thermal's initial depth.
The resulting equations are,
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad\bm{u} =& \\
- \grad \varpi + S_1\hat{z} &
+ \frac{1}{\text{Re}_{\text{ff}}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 =& \\
\frac{1}{\text{Re}_{\text{ff}}}\left(\frac{1}{\text{Pr}_{\text{ff}}\rho_0c_P }\right.&[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\\
&+ \left.\frac{-(\grad_{ad})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right)
\end{split}
\end{gather}
where in this nondimensionalization, the adiabatic temperature gradient is defined
in the next subsection.

\subsection{Atmosphere \& Initial conditions}
We study an ideal gas whose equation of state is $P = \rho T$ and whose stratification
is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z) \\
\rho_0 = T_0^{m_{\text{ad}}},
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and the adiabatic temperature 
gradient in these nondimensional atmospheres is
$$
\grad_{\text{ad}} = -\frac{(e^{n_\rho/m_{ad}} - 1)}{L_z}\frac{m_{ad} + 1}{c_P},
$$
where $L_z$ is the atmospheric depth in units of thermal diameters.

To initialize the simulation, we specify a spherical initial specific entropy perturbation,
\begin{equation}
S_1 = - A(1 - \text{erf}\left(\frac{r' - r_{th}}{\delta}\right))/2,
\label{eqn:thermal_IC}
\end{equation}
where $A = 1$ for our scaled equations.
Here, $r' = \sqrt{r^2 + (z - z_0)^2}$, where $z_0 = L_z - 3*r_{th}$,
with the thermal radius set as $r_{th} = 0.5$, and a smoothing width, $\delta = 0.1$.

\subsection{Fully Compressible Simulations}
In order to verify the validity of our 2D Anelastic simulations, we evolve select
thermals according to the 3D Navier Stokes equations in a cartesian domain. We use
the $(T, \ln\rho)$ formulation of the equations in which we have previously studied fully compressible
convection at low and high Mach number \citep{anders&brown2017, lecoanet&all2014},
\begin{gather}
\frac{D \ln\rho}{Dt} + \DivU = 0 \\
\frac{D \bm{u}}{D t} = -\grad T - T\grad\ln\rho - g\hat{z} + \frac{1}{\rho}\grad\cdot\left(\mu\lilstressT\right) \\
\frac{D T}{Dt} + (\gamma-1)T\DivU  = \frac{1}{\rho c_V}\grad\cdot\left(\kappa \grad T\right) + \frac{\mu}{\rho c_V}\sigma_{ij}\partial_{x_i}u_j.
\end{gather}
where the viscous stress tensor is defined as
\begin{equation}
\sigma_{ij} = \left(\partial_{x_i}u_j + \partial_{x_j}u_i - \frac{2}{3}\delta_{ij}\grad\cdot\bm{u}\right).
\end{equation}
These equations are nondimensionalized on the temperature gradient length scale such that
$\grad_{\text{ad}} = -1$ and the nondimensional timescale is the sound crossing time of that
unit length.

@@

The depth of the atmosphere in these fully compressible domains is set by the number of density
scale heights as $L_z = e^{n_\rho/m_{ad}} - 1$, and in order to compare results from these
simulations to our nondimensionalized thermal simulations, we rescale all length scales in
post-processing by $\ell = 20/L_z$, all timescales by $\tau = \sqrt{\ell \epsilon}$, and
the entropy by $s = \epsilon^{-1}$.

>>>>

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{overview_fig.png}
    \caption{
    \label{fig:overview} }
\end{figure}

@@

\subsection{Evolution of momentum and impulse}
The evolution of thermals as buoyant vortex rings has been well described in
the unstratified, Boussinesq limit for decades (as early as e.g., CITE, and
see \citet{lecoanet&jeevanjee2018} for other sources). Buoyant motions 
in the atmospheres of stars and planets are generally large enough to feel the
atmospheric stratification, and therefore a more thorough treatment of the
evolution of thermals in stratified domains is required to understand 
thermal entrainment in nature. We will use a description of the momentum and
impulse acting on the thermal to develop a theory of depth of thermals as a function
of time, as well as the growth of their radii with depth.

In this work, we study an ideal gas and 
focus on the ideal, low Mach number regime in an adiabatically
stratified atmosphere. In this regime, a linearized equation of state describes
the thermodynamics well, and the fully compressible Euler momentum equation
takes the form \citep{brown&all2012},
\begin{equation}
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} = 
-\grad\varpi - \frac{S_1}{c_P}\bm{g},
\label{eqn:euler_momentum}
\end{equation}
where $\bm{u}$ is the velocity, $\varpi = P_1 / \rho_0$ is the reduced pressure,
$S = c_V\ln T - R\ln\rho$ is the specific entropy, and thermodynamics are broken
down into background (subscript 0) and fluctuating (subscript 1) components.
In this work we find it instructive to examine the full momentum, and so multiplying
this equation by the density, we obtain,
\begin{equation}
\frac{\partial (\rho\bm{u})}{\partial t} + \bm{u}\cdot\grad(\rho\bm{u}) + \rho\bm{u}(\DivU)
= -\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}.
\label{eqn:rho_euler_momentum}
\end{equation}
Hereafter we will define the Lagrangian derivative,
$D/Dt \equiv \partial/\partial t + \bm{u}\cdot\grad$, and we acknowledge that 
the Langrangian derivative commutes with a volume integral such that
\begin{equation*}
\frac{D}{Dt}\int_{\mathcal{V}} f dV = \int_{\mathcal{V}} \left[\frac{Df}{Dt} + f(\DivU)\right]dV.
\end{equation*}
Volume-integrating Eqn.~\ref{eqn:rho_euler_momentum}, we thus find
\begin{equation}
\frac{D\bm{M}}{Dt} = \int_{\mathcal{V}}\left(-\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}\right)dV,
\label{eqn:int_momentum_eqn}
\end{equation}
where the volume-integrated momentum is defined $\bm{M} \equiv \int_{\mathcal{V}}\rho\bm{u} dV$.
At this point we will make the assumption of a plane-parallel atmosphere in which the gravity
is constant, $\bm{g} = -g\hat{z}$, and note that the z-component of the volume-integrated
momentum evolves according to
\begin{equation}
\frac{D M_z}{Dt} = \int_{\mathcal{V}}\left( -\rho\frac{\partial \varpi}{\partial z} + \rho g \frac{S_1}{c_P} \right)dV.
\label{eqn:Mz_definition}
\end{equation}
At this point, we find it useful to define the total buoyancy,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho\, S_1\, \frac{g}{c_P}\, dV,
\label{eqn:tot_buoyancy}
\end{equation}
as in the absence of viscosity and detrainment, this is a conserved quantity during
thermal evolution (REFERENCE FIGURE). Furthermore, in the Boussinesq limit, the
work of \citet{tarshish&all2018} shows that pressure terms reduce the efficacy of
the buoyancy in changing the momentum by effectively reducing the buoyant acceleration
by some factor $\beta \sim 0.5$. As a result, we can approximate the growth of a thermal's
momentum according to
\begin{equation}
\frac{D M_z}{Dt} \approx \beta B.
\label{eqn:theory_momentum}
\end{equation}

After describing the growth of the thermal's integrated momentum with time, we turn
our attention to th ehydrodynamic impulse, which in the stratified limit is defined
\citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector. Impulse is equivalent to the time-integrated
work acting on the fluid resulting in the current fluid motion. 
Per \citet{shivamoggi2010}, changes in the impulse can be expressed as
\begin{equation*}
\frac{D\bm{I}}{D t} = \int_{\mathcal{V}}\frac{\partial(\rho\bm{u})}{\partial t}dV
= B\hat{z} - \int_{\mathcal{V}}\left[\rho\grad\varpi + \Div{\rho\bm{u}\bm{u}}\right]dV ,
\end{equation*}
where, importantly, the \emph{eulerian} derivative of the momentum is inside of the volume
integral here. Under the proper specification of boundary conditions, 
the integral term is zero, and the vertical impulse of a thermal thus straightforwardly
changes in time as
\begin{equation}
\frac{D I_z}{D t} = B.
\label{eqn:change_in_impulse}
\end{equation}
We have thus retrieved the two findings on which our theory of thermal evolution
will be built: \emph{both the impulse and momentum experience constant changes in
time determined by the buoyant nature of the thermal}.

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
    \label{fig:constants} }
\end{figure}

We first note, as in Fig.~\ref{fig:constants}a, that the buoyancy in our thermals
is not necessarily perfectly constant in time, and that effects like detrainment result
in a minor reduction of the buoyant signature of our thermals. We thus express the
buoyancy in terms of two constants,
\begin{equation}
B \approx \chi B_0,
\end{equation}
where $B_0$ is the initial buoyancy of the thermal, and $\chi$ is a constant of
O(1) which represents this detrainment. We then integrating 
Eqn.~\ref{eqn:change_in_impulse} in time,
\begin{equation*}
I_z = \chi B_0 t + I_0.
\end{equation*}
In the low-Mach number limit in which changes in density from the background are negligible,
the impulse of a vortex ring can be approximated as
\begin{equation*}
I_z \approx \pi \rho r^2 \Gamma_0,
\end{equation*}
where $r$ is the radius of the thermal from its axis
of symmetry to the maxima of its buoyant signature, and
$\Gamma_0 = \int_{\mathcal{A}} (\grad\times\bm{u})dA$ is the integrated circulation
in a cross-section of the vortex ring. We take $\Gamma_0$ to be constant, which is
a decent assumption in our experiments, see Fig.~\ref{fig:constants}b.
Combining these two expressions, we retrieve our first result,
\begin{equation}
r = \sqrt{\frac{\chi B_0 t + I_0}{\pi\rho\Gamma_0}}.
\label{eqn:r_theory}
\end{equation}
In the Boussinesq limit where $\rho \rightarrow \text{constant}$, 
we retrieve the $r \propto \sqrt{t}$ scaling found in the Boussinesq regime by
\citet{lecoanet&jeevanjee2018}. We find that the inclusion of stratification adds the
additional complexity of $r \propto \rho^{-1/2}$, such that downward-propogating 
vortex rings (as studied here) will entrain less than boussinesq thermals,
and upward-propogating rings will entrain more. In the absence of buoyancy ($B = 0$),
this result aligns with the prediction for purely horizontal compression noted by
\citet{brandenburg2016} of $r^2 \propto \rho^{-1}$.

The momentum can likewise be integrated like the impulse,
\begin{equation*}
M_z = \beta\chi B_0 t + M_0.
\end{equation*}
For the proper choice of vertical velocity, $w_{\text{th}}$, and volume, $\mathcal{V}$, the
momentum can be expressed precisely as 
\begin{equation*}
M_z = \rho \mathcal{V} w_{\text{th}},
\end{equation*}
and the volume can be approximated as $\mathcal{V} = f r^3$, where $f$ is a parameter
which we take to be constant. This assumption is perhaps not as perfect as the
assumptions of constant $B_0$ or $\Gamma_0$ (see Fig.~\ref{fig:constants}c), but
it is nevertheless not a bad assumption. Combining our approximate expressions,
and inserting our theoretical description of $r$ (Eqn.~\ref{eqn:r_theory}),
we retrieve
\begin{equation}
\rho^{-1/2} w_{\text{th}} = \left(\frac{(\pi \Gamma)^{3/2}}{f}\right)\frac{\beta\chi B_0 t + M_0}{(\chi B_0 t + I_0)^{3/2}}.
\end{equation}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, and making the
assumption that the vortex ring starts at a ``virtual origin'' (CITE) at $t = 0$ where
it has no impulse or momentum ($I_0 = M_0 = 0$), an integrable expression for the evolution
of thermal position over time can be retrieved,
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{\beta(\pi\Gamma)^{3/2}}{f(\chi B_0)^{1/2}}\right)\frac{dt}{t^{1/2}}
\label{eqn:dz_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this
result can be integrated with $\rho(z_{\text{th}})$ plugged in in order to
find the position of the thermal as a function of time. We leave this result
general for now, and will integrate it using the polytropic stratification used in
our simulations at the end of section \ref{sec:experiment}.

@@

\subsection{Anelastic Simulations}
In this work, we primarily study the evolution of 2D, azimuthally symmetric, anelastic
thermals in cylindrical coordinates. 
We later verify that select simulations produce the same results as 3D fully compressible
simulations in cartesian domains (see sec. REF). 
The LBR anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\td{\grad}\cdot\td{\bm{u}} = -\td{w}\partial_{\tilde{z}} \ln\rho_0 
\label{eqn:AN_continuity_full}\\
\frac{D \td{\bm{u}}}{D \td{t}} = -\td{\grad} \td{\varpi} + g\frac{\td{S_1}}{c_P}\hat{z} + \frac{1}{\rho_0}\td{\grad}\cdot\left(\mu\td{\lilstressT}\right)
\label{eqn:AN_momentum_full}\\
\frac{D \td{S_1}}{D\td{t}} = \frac{1}{\rho c_P}\td{\grad}\cdot\left(\kappa T_0 \td{\grad} \td{S_1}\right) + \frac{\mu}{\rho_0 T_0}\td{\sigma_{ij}}\partial_{\td{x_i}}\td{u_j}
\label{eqn:AN_energy_full}.
\end{gather}
where terms with tildes are dimensional, and where
$D/D\td{t} = \partial/\partial \td{t} + \td{\bm{u}}\cdot\td{\grad}$. In our azimuthally symetric
domain, we assume that $\partial_\phi = 0$; as the initial conditions of our simulations are at rest and have
no azimuthal velocity, $u_\phi$, we explicitly impose that $u_\phi = 0$; therefore $\bm{u} = u_r \hat{r} + w\hat{z}$. 
Under this approximation, the components of the stress tensor are
\begin{equation}
\begin{split}
&\td{\sigma}_{rr} = 2\frac{\partial \td{u_r}}{\partial \td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,
\td{\sigma}_{\phi\phi} = 2\frac{\td{u_r}}{\td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}}, \\
&\td{\sigma}_{zz}       = 2\frac{\partial \td{w}}{\partial \td{z}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,\,
\td{\sigma}_{rz}     = \td{\sigma}_{zr} = \frac{\partial \td{w}}{\partial \td{r}} + \frac{\partial \td{u_r}}{\partial \td{z}}, \\
&\td{\sigma}_{r\phi}  = \td{\sigma}_{\phi r}  
 = \td{\sigma}_{\phi z} = \td{\sigma}_{z \phi}  = 0,\qquad
\end{split}
\end{equation}
Furthermore, we assume that the dynamic viscosity, $\mu = \rho_0 \nu$, and the
thermal conductivity, $\kappa = \rho_0 \chi$, are both uniform throughout the domain and
constant in time.
The diffusivities $\nu$ and $\chi$ therefore scale inversely with the density.

We nondimensionalize these equations in the same manner as in
\citet{lecoanet&jeevanjee2018} such that
the length scale is the diameter of the initial thermal perturbation
and the velocity scale is the freefall velocity. The timescale is thus
the freefall crossing time of this unit length. Mathematically,
\begin{equation}
\begin{split}
\td{\grad}\rightarrow(\td{L}_{th}^{-1})\grad, \qquad&
\td{S}_1 \rightarrow(\Delta\td{S})S_1,\\
\td{\bm{u}} \rightarrow (\td{u}_{th})\bm{u}, \qquad&
\td{\varpi} \rightarrow (\td{u}_{th}^2)\varpi,\\
\partial_{\tilde{t}} \rightarrow (\td{u}_{th}/\td{L}_{th})\partial_t,\qquad&
\end{split}
\end{equation}
with
\begin{equation}
\tilde{u}_{th}^2 = \frac{g \tilde{L}_{th} \Delta \tilde{s}}{c_P}, \qquad
\text{Re}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\nu}, \qquad
\text{Pr}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\chi}.
\end{equation}
As the diffusivities scale with depth, Re$_{\text{ff}}$ is specified at the
thermal's initial depth.
The resulting equations are,
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad&\bm{u} = \\
- \grad \varpi + S_1\hat{z} &
+ \frac{1}{\text{Re}_{\text{ff}}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 =& \\
\frac{1}{\text{Re}_{\text{ff}}}\left(\frac{1}{\text{Pr}_{\text{ff}}\rho_0c_P }\right.&[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\\
&+ \left.\frac{-(\grad_{\text{ad}})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right),
\end{split}
\end{gather}
where $\grad_{\text{ad}} \equiv \tilde{L}_{\text{th}} \frac{g}{c_P}$.

\subsection{Atmosphere \& Initial conditions}
We study an ideal gas whose equation of state is $P = \rho T$ and whose stratification
is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z) \\
\rho_0 = T_0^{\,m_{\text{ad}}},
\label{eqn:polytrope}
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and the adiabatic temperature 
gradient in these nondimensional atmospheres is set with $g = m_{\text{ad}} + 1$
and $\tilde{L}_{\text{th}} = (e^{n_\rho/m_{\text{ad}}} - 1)/L_z$, where
$n_\rho$ is the number of density scale heights spanned by the atmosphere
and $L_z = 20$ is the nondimensional depth of the atmosphere in units of
thermal diameters.

To initialize the simulation, we specify a spherical initial specific entropy perturbation,
\begin{equation}
S_1 = - \frac{A}{2}\left[1 - \text{erf}\left(\frac{r' - r_{th}}{\delta}\right)\right],
\label{eqn:thermal_IC}
\end{equation}
where $A = 1$ for our scaled equations.
Here, $r' = \sqrt{r^2 + (z - z_0)^2}$, where $z_0 = L_z - 3r_{\text{th}}$,
with the thermal radius set as $r_{th} = 0.5$, and a smoothing width, $\delta = 0.1$.

\subsection{Fully Compressible Simulations}
In order to verify the validity of our 2D Anelastic simulations, we evolve select
thermals according to the 3D Navier Stokes equations in a cartesian domain. We use
the $(T, \ln\rho)$ formulation of the equations in which we have previously studied fully compressible
convection at low and high Mach number \citep{lecoanet&all2014, anders&brown2017},
\begin{gather}
\frac{D \ln\rho}{Dt} + \DivU = 0 \\
\frac{D \bm{u}}{D t} = -\grad T - T\grad\ln\rho - g\hat{z} + \frac{1}{\rho}\grad\cdot\left(\mu\lilstressT\right) \\
\frac{D T}{Dt} + (\gamma-1)T\DivU  = \frac{1}{\rho c_V}\grad\cdot\left(\kappa \grad T\right) + \frac{\mu}{\rho c_V}\sigma_{ij}\partial_{x_i}u_j.
\end{gather}
where the viscous stress tensor is defined as
\begin{equation}
\sigma_{ij} = \left(\partial_{x_i}u_j + \partial_{x_j}u_i - \frac{2}{3}\delta_{ij}\grad\cdot\bm{u}\right).
\end{equation}
These equations are nondimensionalized on the temperature gradient length scale such that
$\grad_{\text{ad}} = -1$ and the nondimensional timescale is the sound crossing time of that
unit length at the top of the domain.

@@

The depth of the atmosphere in these fully compressible domains is set to $L_z = e^{n_\rho/m_{ad}} - 1$.
In order to compare results from these
simulations to our Anelastic thermal simulations, we rescale all outputs in post-processing
by a length scale factor of $\ell = 20/L_z$, a timescales factor of  $\tau = \sqrt{\ell \epsilon}$,
and a specific entropy scaling of $s = \epsilon^{-1}$.

\subsection{Numerics}
We evolve our simulations forward in time using the 
Dedalus\footnote{\url{http://dedalus-project.org/}} 
pseudospectral framework \citep{burns&all2016} to time-evolve
our equations. For our 2D simulations, we use 
using an implicit-explicit (IMEX), third-order, four-stage
Runge-Kutta timestepping scheme RK443 \citep{ascher&all1997},
and for our 3D simulations we use the second order
semi-implicit backward differentiation formulation
SBDF2 \citep{wang&ruuth2008}.

Our 3D simulations are decomponsed on Fourier bases in the horizontal
directions ($x, y \in [-L_r, L_r]$) and Chebyshev bases vertically
($z \in [0, L_z]$) with impenetrable, stress free, fixed-temperature
boundary conditions at the upper and lower boundaries 
($T_1 = w = \partial_z u = \partial_z v = 0$ at $z = [0, L_z]$).
Our 2D simulations are decomposed on a Fourier $(z \in [0, L_z])$ and
Chebyshev $(r \in [0, L_r])$ domain, with boundary conditions of
$\partial_r S_1 = w = (\grad\times\bm{u})_{\phi} = 0$ at $r = L_r$.
